---
title: "Dplyr"
author: "Viv Tulloch"
date: "6/9/2019"
output: html_document
---

shortcut for ``` cmd/ctrl+alt+i
run = cmd + entr


loading tidyverse

```{r}
library(tidyverse)

```

if tidyverse doesn't work load ddply, readr, gglpot2

Load data
```{r}
# loading the gapminder dataset from github

gapminder <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")


```


```{r}

head(gapminder)
summary(gapminder)
tail(gapminder)
str(gapminder)  #structure
names(gapminder)

```

Look at individual dataframe variables

```{r}

#show me the lifeExp column
gapminder$lifeExp
gapminder$continent

```


Ues ddplyr to 'filter()' or subset data row-wise

```{r filter subset}

filter(gapminder,lifeExp < 29)

#just look at mexico

filter(gapminder, country == Mexico, year==2002)

brz <- filter(gapminder, country == Brazil, year >= 1987 & year <=2007)
mean(brz$lifeExp)

```

Learning the 'select()' function

```{r}
select(gapminder, year, country, lifeExp)
```

