---
title: "Dplyr"
author: "Viv Tulloch"
date: "6/9/2019"
output: html_document
---

shortcut for ``` cmd/ctrl+alt+i
run = cmd + entr
<- alt - 

loading tidyverse

```{r}
library(tidyverse)

```

if tidyverse doesn't work load ddply, readr, gglpot2

Load data
```{r}
# loading the gapminder dataset from github

gapminder <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv")


```


```{r}

head(gapminder)
summary(gapminder)
tail(gapminder)
str(gapminder)  #structure
names(gapminder)

```

Look at individual dataframe variables

```{r}

#show me the lifeExp column
gapminder$lifeExp
gapminder$continent

```


Ues ddplyr to 'filter()' or subset data row-wise

```{r filter subset}

filter(gapminder,lifeExp < 29)

#just look at mexico

filter(gapminder, country == Mexico, year==2002)

brz <- filter(gapminder, country == Brazil, year >= 1987 & year <=2007)
mean(brz$lifeExp)

```

Learning the 'select()' function

```{r}
#these 2 lines do the same thing - first one selects columns, second drops columns
select(gapminder, year, country, lifeExp)
select(gapminder, -continent, -gdpPercap, -pop)
```


Using 'filter()' and 'select()' together

```{r}
gap_cambodia <- filter(gapminder, country = "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)
```

Pipe Operator

Learning the pipe %>%
ctrl + shift + M

```{r}
gapminder %>% head()
head(gapminder)

# create a new object first filtering for cambodia AND THEN dropping columns
gap_cambodia <- filter(gapminder, country = "Cambodia") %>% 
  select(-continent, -lifeExp)


gap_mexico <- gapminder %>% 
  filter(country = "Cambodia") %>%  # AND THEN
  select(-continent, -lifeExp)


```

Learn 'mutate()' wich adds a new column
then you tell R what to put in that column

```{r}
gapminder %>% 
  mutate(gdp = pop * gdpPercap)  ## gdp is the first columns name

names(gapminder)

# get asian countries and pop in thousands
gapminder %>% 
  filter(continent = "Asia", year == 2007) %>% 
  mutate(pop_thousands = pop/1000) %>% 
  select(country, year, pop_thousands)

mean(df$pop_thousands)
```

Learn the 'group_by()' function

```{r}
gapminder %>% 
  filter(year == 2002) %>% 
  group_by(continent) %>% 
  mutate(cont_pop = sum(pop))

```

Add in 'summarize()'

```{r}

gapminder %>% 
  filter(year == 2002) %>% 
  group_by(continent) %>% 
  summarize(cont_pop = sum(pop)) #summarize by continent
```

Often good to ungroup after grouping data

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
   mutate(cont_pop = sum(pop))
  ungroup()
```


'arrange()'

```{r}
gapminder %>% 
  group_by(continent,year) %>% 
  summarize(cont_pop = sum(pop)) %>% 
  arrange(year)

# what is max gdp per continent

gapminder %>% 
  mutate(gdp = pop * gdpPercap) %>% # create new column with gdp
  group_by(continent) %>% 
  mutate(max_gdp = max(gdp)) %>% # create new column with max gdp
  filter(gdp == max_gdp)

  
```

